Associations between Chlorophyll \emph{a} and various Microcystin-LR Health Advisory Concentrations

### 2015-09-25 manuscript/manuscript.tex
rendered Betty's commmetns

Cyanobacteria harmful algal blooms (cHABs) are associated with a wide
range of adverse health effects that stem mostly from the presence of
cyanotoxins. To help protect against these impacts, several health
advisory levels have been set for some toxins. In particular, one of the
more common toxins, microcystin, has several advisory levels set for
drinking water and recreational use and managing water bodies to meet
those levels could have far reaching benefits. However, compared to
other water quality measures, measurements of microcystin are not common
and current field measurement techniques have limited precision and
accuracy. Addressing these issues will take time and resources. Thus,
there is utility in finding indicators of microcystin that are already
widely available, can be estimated quickly and \emph{in situ}, and used
as a first defense against high levels of microcystin. In particular,
chlorophyll \emph{a} is very commonly measured, can be estimated
\emph{in situ}, and has been shown to be positively associated with
microcystin. In this paper we use this association to provide estimates
of chlorophyll \emph{a} that if exceeded would be indicative of a higher
probability of exceeding select health advisory concentrations for
microcystin-LR. Using the 2007 National Lakes Assessment and a
conditional probability approach that has been used in other water
quality settings, we identify chlorophyll \emph{a} concentrations that
are more likely than not to be associated with an exceedance of a
microcystin health advisory level. We look at the recent US EPA
standards for drinking water as well as the World Health Organization
levels for drinking water and recreational use. For microcystin
concentrations of 0.3, 1, 1.6, and 2 we find chlorophyll \emph{a}
concentrations of 23.68, 63.94, 89.71, and 102.6, respectively. When
managing for these various microcystin levels exceeding these reported
chlorophyll \emph{a} concentrations should be a trigger for further
testing and possible management action.

@@

In the summer of 2014, the city of Toledo, OH was forced to shut down
their municipal water supply due in part to an excess of microcystin-LR
that resulted from a ongoing cyanobacteria harmful algal bloom (cHAB) in
Lake Erie {[}REFS{]}. Since this event, significant legislation has been
passed in the United States and the US Environmental Protection Agency
(USEPA) has released suggested microcystin-LR concentrations that would
trigger health advisories. MORE ON THE LEVELS. While these levels and
associated advisories are likely to help mitigate the impacts from
harmful algal blooms, they are not without complications.

One of these complications is that they rely on available measurements
of microcystin-LR. This toxin can be measured in the field using test
strips but these are a coarse measure at best and currently available
test strips focus on 1 and 10 \(\mu\)g/L {[}REFS{]}. Measurements with
greater accuracy and precision require taking water samples and
processing those in a lab to determine the toxin concentration
{[}REFS{]}. Additionally, microcystin-LR is, currently, not a routinely
collected water quality parameter,thus, availability of microcystin-LR
data may limit our ability to screen water bodies for exceedances of the
various health advisories concentrations. Until microcystin-LR
concentrations are more widely collected an alternative measure is
needed. Fortunately, microcystin-LR has been shown to be associated with
several other, more easily measured components of water quality.

Chlorophyll \emph{a} is a very commonly measured components of water
quality that is also known to be associated with Microsystin-LR
concentrations {[}REFS{]}. Additionally there are many rapid
measurements for assessing chlorophyll \emph{a} levels \emph{in situ}.
For instance, there are small or hand held flourometers that provide
reliable measurements {[}REFS{]}. Given these facts, it might be
possible to identify chlorophyll \emph{a} concentrations that would be
associated with the various Microcystin-LR health advisory levels.
Identifying these associations would provide another reliable tool for
water resource managers to use to help manage the threat to public
health posed by cHABs and would be especially useful in the absence of
microcystin-LR concentrations. Thus, the goal of this paper is to utlize
the National Lakes Assessment data and identify chlorophyll \emph{a}
concentrations that are associated with higher probabilities of
exceeding several microcystin-LR health advisory concentrations {[}NLA
REF{]}. So that others may repeat or adjust this analysis, the data,
code, and this manuscript are freely available via
\href{https://github.com/USAPE/microcystinchla}{\url{https://github.com/USAPE/microcystinchla}}.

@@

@@

@@

@@

>>>>

Cyanobacteria harmful algal blooms (cHABs) are associated with a wide
range of adverse health effects that stem mostly from the presence of
cyanotoxins. To help protect against these impacts, several health
advisory levels have been set for some toxins. In particular, one of the
more common toxins, microcystin, has several advisory levels set for
drinking water and recreational use. However, compared to other water
quality measures, field measurements of microcystin are not commonly
available due to cost and require advanced understanding to interpret
results. Addressing these issues will take time and resources. Thus,
there is utility in finding indicators of microcystin that are already
widely available, can be estimated quickly and \emph{in situ}, and used
as a first defense against high levels of microcystin. In particular,
chlorophyll \emph{a} is very commonly measured, can be estimated
\emph{in situ}, and has been shown to be positively associated with
microcystin. In this paper we use this association to provide estimates
of chlorophyll \emph{a} that if exceeded would be indicative of a higher
probability of exceeding select health advisory concentrations for
microcystin-LR. Using the 2007 National Lakes Assessment and a
conditional probability approach that has been used in other water
quality settings, we identify chlorophyll \emph{a} concentrations that
are more likely than not to be associated with an exceedance of a
microcystin health advisory level. We look at the recent US EPA
standards for drinking water as well as the World Health Organization
levels for drinking water and recreational use. For microcystin
concentrations of 0.3, 1, 1.6, and 2 we find chlorophyll \emph{a}
concentrations of 23.68, 63.94, 89.71, and 102.6, respectively. When
managing for these various microcystin levels exceeding these reported
chlorophyll \emph{a} concentrations should be a trigger for further
testing and possible management action.

@@

In the summer of 2014, the city of Toledo, OH was forced to shut down
their municipal water supply due in part to an excess of microcystin-LR
that resulted from a ongoing cyanobacteria harmful algal bloom (cHAB) in
Lake Erie (Rinta-Kanto et al. 2009, Jetoo et al. 2015). Since this
event, significant legislation has been passed in the United States and
the US Environmental Protection Agency (USEPA) has released suggested
microcystin-LR (one of the more common toxins) concentrations that would
trigger health advisories (McElhiney and Lawton 2005, Zurawell et al.
2005). MORE ON THE LEVELS. While these levels and associated advisories
are likely to help mitigate the impacts from harmful algal blooms, they
are not without complications.

One of these complications is that they rely on available measurements
of microcystin-LR. While laboratory testing remains the gold standard
for quantifying microcystin-LY concentrations in water samples, several
field test kits have been developed. Even though field tests provide a
much needed means for rapid assessment, they are infrequently available
to lake managers. These kits are moderately expensive (approximately
\$150-\$200 depending on specific kit) with a limited shelf life
(typically one year) (James et al. 2011, Aranda-Rodriguez et al. 2015).
Additionally, each technique requires nuanced understanding of the
detection method (e.g., limit of detection, specific microcystin
variants being measured, and sampling protocol). Fortunately,
microcystin-LR has been shown to be associated with several other, more
commonly measured and understood components of water quality.

Chlorophyll \emph{a} is a very commonly measured components of water
quality that is also known to be associated with Microsystin-LR
concentrations (Lee et al. 2000, Paerl and Otten (2013)). Additionally
there are many rapid measurements for assessing chlorophyll \emph{a}
levels \emph{in situ}. For instance, there are small or hand held
flourometers that provide reliable measurements {[}REFS{]}. Given these
facts, it might be possible to identify chlorophyll \emph{a}
concentrations that would be associated with the various Microcystin-LR
health advisory levels. Identifying these associations would provide
another reliable tool for water resource managers to use to help manage
the threat to public health posed by cHABs and would be especially
useful in the absence of microcystin-LR concentrations. Thus, the goal
of this paper is to utlize the National Lakes Assessment data and
identify chlorophyll \emph{a} concentrations that are associated with
higher probabilities of exceeding several microcystin-LR health advisory
concentrations {[}NLA REF{]}. So that others may repeat or adjust this
analysis, the data, code, and this manuscript are freely available via
\href{https://github.com/USAPE/microcystinchla}{\url{https://github.com/USAPE/microcystinchla}}.

@@

\begin{verbatim}
## Loading required package: grid
\end{verbatim}

@@

Aranda-Rodriguez, R., Z. Jin, J. Harvie, and A. Cabecinha. 2015.
Evaluation of three field test kits to detect microcystins from a public
health perspective. Harmful Algae 42:34--42.

James, R., A. Gregg, A. Dindal, and J. McKernan. 2011. Environmental
technology verification report: Abraxis microcystin test kits. Online
document. Accessed online: June 22.

Jetoo, S., V. I. Grover, and G. Krantzberg. 2015. The toledo drinking
water advisory: Suggested application of the water safety planning
approach. Sustainability 7:9787--9808.

Lee, S., M.-H. Jang, H.-S. Kim, B.-D. Yoon, and H.-M. Oh. 2000.
Variation of microcystin content of microcystis aeruginosa relative to
medium n: P ratio and growth stage. Journal of Applied Microbiology
89:323--329.

McElhiney, J., and L. A. Lawton. 2005. Detection of the cyanobacterial
hepatotoxins microcystins. Toxicology and Applied Pharmacology
203:219--230.

@@

Paerl, H. W., and T. G. Otten. 2013. Harmful cyanobacterial blooms:
Causes, consequences, and controls. Microbial ecology 65:995--1010.

@@

Rinta-Kanto, J. M., E. A. Konopko, J. M. DeBruyn, R. A. Bourbonniere, G.
L. Boyer, and S. W. Wilhelm. 2009. Lake erie microcystis: Relationship
between microcystin production, dynamics of genotypes and environmental
parameters in a large lake. Harmful Algae 8:665--673.

Zurawell, R. W., H. Chen, J. M. Burke, and E. E. Prepas. 2005.
Hepatotoxic cyanobacteria: A review of the biological importance of
microcystins in freshwater environments. Journal of Toxicology and
Environmental Health, Part B 8:1--37.
